Bird Extinction

### 2018-09-14 bamm_extinction/BirdExtinction.Rmd
Long time no update

```{r}
library(ape)
library(phytools)
library(dplyr)
library(coda)
library(ggplot2)
library(ggridges)
library(phytools)
library(BAMMtools)
library(diversitree)
source("functions/R_test_BAMM_functions.R")
source("functions/check_and_fix_ultrametric.R")
# bird.tree.eric <- read.nexus('data/ErisonGeneticFull.nex')
# averageTree(bird.tree.eric)
```

@@

##BAMM Setup

@@

``bamm -c control_consensus_eric.txt``

@@

###MCC Tree

```{r}
is.ultrametric(mcc.trees.eric[["tree_2241"]])
is.binary.tree(mcc.trees.eric[["tree_2241"]])
# Now to check min branch length:
min(mcc.trees.eric[["tree_2241"]]$edge.length)
```

Create a control file with same parameters as the consensus tree:

```{r}
priors <- setBAMMpriors(mcc.trees.eric[["tree_2241"]], outfile = NULL)
write.tree(mcc.trees.eric[["tree_2241"]], 'mcc_eric_2241.tre')

# Here is a block of parameters for the control file
params = list(treefile = 'mcc_eric_2241.tre',
            sampleFromPriorOnly = 0,
            runMCMC = 1,
            initializeModel = 1,
            useGlobalSamplingProbability = 1,
            globalSamplingFraction = 0.67,
            updateLambdaInitScale = 2.0,
            updateMuInitScale = 2.0,
            updateLambdaShiftScale = 0.05,
            updateEventRateScale = 4.0,
            localGlobalMoveRatio = 10.0,
            modeltype = 'speciationextinction', #rather than trait

bammcontrolfile <- paste("control_", "mcc_eric_2241", ".txt", sep="")

>>>>

#Setup

```{r, warning = FALSE}
library(ape)
library(phytools)
library(dplyr)
library(coda)
library(ggplot2)
library(ggridges)
library(phytools)
library(BAMMtools)
library(diversitree)
source("functions/R_test_BAMM_functions.R")
source("functions/check_and_fix_ultrametric.R")
# bird.tree.eric <- read.nexus('data/ErisonGeneticFull.nex')
# averageTree(bird.tree.eric)
```

@@

@@

#BAMM Setup

##Load trees and check them

Read in 100 Passerine Trees from birdtree.org (5966 species) based on the Hackett backbone.
```{r}
passerine.trees <- read.nexus('data/trees/Passerine_Trees_Full.nex')
```

For BAMM to work, we need to check to see if trees are ultrametric, binary and edge lengths are greater than 0.

We can check this on the first tree
```{r}
is.ultrametric(passerine.trees[["tree_8950"]])
is.binary.tree(passerine.trees[["tree_8950"]])
# Now to check min branch length:
min(passerine.trees[["tree_8950"]]$edge.length)
```

> good

We can check all 100 trees with this slightly messy script:
```{r, eval=FALSE}
is.ultrametric(passerine.trees)
is.binary.tree(passerine.trees)
# Now to check min branch length:
name.passerine.tree <- names(passerine.trees)
for(x in name.passerine.tree) {
  if (min(passerine.trees$edge.length[[x]]) > 0)
    print("good")

  else print("bad")
}
```

> Good, we can use any tree for BAMM

##Writing the Control Files

Create a control file for all the trees (100 files and 100 trees). To do this we can use the ``sapply`` function.

```{r, results = 'hide'}
name.passerine.tree <- names(passerine.trees)

priors <- sapply(name.passerine.tree, function(x) {
  setBAMMpriors(passerine.trees[[x]], outfile = NULL)
})

sapply(name.passerine.tree, function(x) {
  write.tree(passerine.trees[[x]], paste("data/bamm_files/", x, ".tre", sep=""))
})

# Here is a block of parameters for the control file. We can make a control file for each tree:
params <- list()
for (x in name.passerine.tree) {

# GENERAL SETUP AND DATA INPUT

params[[x]] <- list(modeltype = 'speciationextinction',
# Specify "speciationextinction" or "trait" analysis

treefile = paste(x, ".tre", sep=""),
# File name of the phylogenetic tree to be analyzed

runInfoFilename = 'run_info.txt',
# File name to output general information about this run

sampleFromPriorOnly = 0,
# Whether to perform analysis sampling from prior only (no likelihoods computed)

runMCMC = 1,
# Whether to perform the MCMC simulation. If runMCMC = 0, the program will only
# check whether the data file can be read and the initial likelihood computed

loadEventData = 0,                       
# Whether to load a previous event data file

eventDataInfile = 'event_data_in.txt',
# File name of the event data file to load, used only if loadEventData = 1

initializeModel = 1,
# Whether to initialize (but not run) the MCMC. If initializeModel = 0, the
# program will only ensure that the data files (e.g., treefile) can be read

useGlobalSamplingProbability = 1,
# Whether to use a "global" sampling probability. If False (0), expects a file
# name for species-specific sampling probabilities (see sampleProbsFilename)

globalSamplingFraction = 1,
# The sampling probability. If useGlobalSamplingProbability = 0, this is ignored
# and BAMM looks for a file name with species-specific sampling fractions

sampleProbsFilename = 'sample_probs.txt',
# File name containing species-specific sampling fractions

seed = as.numeric(gsub("tree_", "", x, perl = TRUE)),
# Seed for the random number generator. Set for reproducibility to the number of the treefile

overwrite = 1,
# If True (1), the program will overwrite any output files in the current
# directory (if present)

# PRIORS

expectedNumberOfShifts = 100,
# prior on the number of shifts in diversification
# Suggested values: 
#     expectedNumberOfShifts = 1.0 for small trees (< 500 tips)
#  expectedNumberOfShifts = 10 or even 50 for large trees (> 5000 tips)

lambdaInitPrior = as.numeric(priors['lambdaInitPrior', x]),
# Prior (rate parameter of exponential) on the initial lambda value for rate
# regimes

lambdaShiftPrior = 0.05,
# Prior (std dev of normal) on lambda shift parameter for rate regimes
# You cannot adjust the mean of this distribution (fixed at zero, which is
# equal to a constant rate diversification process)

muInitPrior = as.numeric(priors['muInitPrior', x]),
# Prior (rate parameter of exponential) on extinction rates

lambdaIsTimeVariablePrior = 1,
# Prior (probability) of the time mode being time-variable (vs. time-constant)

# MCMC SIMULATION SETTINGS & OUTPUT OPTIONS

numberOfGenerations = '100000000',
# Number of generations to perform MCMC simulation

mcmcOutfile = 'mcmc_out.txt',
# File name for the MCMC output, which only includes summary information about
# MCMC simulation (e.g., log-likelihoods, log-prior, number of processes)

mcmcWriteFreq = 1000,
# Frequency in which to write the MCMC output to a file

eventDataOutfile = 'event_data.txt',
# The raw event data (these are the main results). ALL of the results are
# contained in this file, and all branch-specific speciation rates, shift
# positions, marginal distributions etc can be reconstructed from this output.
# See R package BAMMtools for working with this output

eventDataWriteFreq = 1000,
# Frequency in which to write the event data to a file

printFreq = 10000,
# Frequency in which to print MCMC status to the screen

acceptanceResetFreq = 1000,
# Frequency in which to reset the acceptance rate calculation
# The acceptance rate is output to both the MCMC data file and the screen

outName = x,
# Optional name that will be prefixed on all output files (separated with "_")
# If commented out, no prefix will be used

# OPERATORS: MCMC SCALING OPERATORS

updateLambdaInitScale = 2,
# Scale parameter for updating the initial speciation rate for each process

updateLambdaShiftScale = 0.1,
# Scale parameter for the exponential change parameter for speciation

updateMuInitScale = 2,
# Scale parameter for updating initial extinction rate for each process

updateEventLocationScale = 0.1,
# Scale parameter for updating LOCAL moves of events on the tree
# This defines the width of the sliding window proposal

updateEventRateScale = 4,
# Scale parameter (proportional shrinking/expanding) for updating
# the rate parameter of the Poisson process

# OPERATORS: MCMC MOVE FREQUENCIES

updateRateEventNumber = 1,
# Relative frequency of MCMC moves that change the number of events

updateRateEventPosition = 0.25,
# Relative frequency of MCMC moves that change the location of an event on the
# tree

updateRateEventRate = 1,
# Relative frequency of MCMC moves that change the rate at which events occur

updateRateLambda0 = 1,
# Relative frequency of MCMC moves that change the initial speciation rate
# associated with an event

updateRateLambdaShift = 1,
# Relative frequency of MCMC moves that change the exponential shift parameter
# of the speciation rate associated with an event

updateRateMu0 = 1,
# Relative frequency of MCMC moves that change the extinction rate for a given
# event

updateRateLambdaTimeMode = 0,
# Relative frequency of MCMC moves that flip the time mode
# (time-constant <=> time-variable)

localGlobalMoveRatio = 10,
# Ratio of local to global moves of events

# INITIAL PARAMETER VALUES

lambdaInit0 = 0.032,
# Initial speciation rate (at the root of the tree)

lambdaShift0 = 0,
# Initial shift parameter for the root process

muInit0 = 0.005,
# Initial value of extinction (at the root)

initialNumberEvents = 0,
# Initial number of non-root processes

# METROPOLIS COUPLED MCMC

numberOfChains = 1,
# Number of Markov chains to run

deltaT = 0.01,
# Temperature increment parameter. This value should be > 0
# The temperature for the i-th chain is computed as 1 / [1 + deltaT * (i - 1)]

swapPeriod = 1000,
# Number of generations in which to propose a chain swap

chainSwapFileName = 'chain_swap.txt',
# File name in which to output data about each chain swap proposal.
# The format of each line is [generation],[rank_1],[rank_2],[swap_accepted]
# where [generation] is the generation in which the swap proposal was made,
# [rank_1] and [rank_2] are the chains that were chosen, and [swap_accepted] is
# whether the swap was made. The cold chain has a rank of 1.

# NUMERICAL AND OTHER PARAMETERS

minCladeSizeForShift = 3,
# Allows you to constrain location of possible rate-change events to occur
# only on branches with at least this many descendant tips. A value of 1
# allows shifts to occur on all branches.

segLength = 0.025,
# Controls the "grain" of the likelihood calculations. Approximates the
# continuous-time change in diversification rates by breaking each branch into
# a constant-rate diversification segments, with each segment given a length
# determined by segLength. segLength is in units of the root-to-tip distance of
# the tree. So, if the segLength parameter is 0.01, and the crown age of your
# tree is 50, the "step size" of the constant rate approximation will be 0.5.
# If the value is greater than the branch length (e.g., you have a branch of
# length < 0.5 in the preceding example) BAMM will not break the branch into
# segments but use the mean rate across the entire branch.

outName = x)
  }

bammcontrolfile <- list()
for (x in name.passerine.tree) {
  bammcontrolfile[x] <- paste("data/bamm_files/control_", x, ".txt", sep="")
}

# Now writing control parameters to file

for (x in name.passerine.tree) {generateControlFile(file = bammcontrolfile[[x]], type = "diversification", params = params[[x]])}
```

BAMM can be run through the terminal through the following syntax: ``bamm -c control_tree_xxxx.txt``. To generate these commands we can use a loop function, from which we get:

```{r}
bamm.commands <- list()
for (x in name.passerine.tree) {
  bamm.commands[x] <- paste("bamm -c control_", x, ".txt", sep="")
}
```

@@
